Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-04-09 doc/manuscript_hkk_drought_sensitivity.Rmd
update with figure numbers as vars

## Figures
droughts = 1
drought_responses = 2

@@

![***Figure `r droughts`. Climatic characteristics of drought years and average years*** in the Huai Kha Khaeng ForestGEO plot. For four daily climate variables dry days, precipitation, maximum temperature and VPD, panel a) represents rolling means for 30 days around the focal day for that year. Grey line represents the mean of this rolling mean from 2008 to 2019 and ribbons represent standard deviation around this mean. Panel b) represents anomalies of this rolling mean from the values across the timeseries. Values, therefore represents the number of standard deviations from the mean on a given day. c) shows Standardised Precipitation Evapotranspiration Index (SPEI) values for each month in 2010 and 2015. Across all panels, light blue background represents the average wet season, from May to October. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1.png)

@@

![***Figure `r drought_responses`: Growth increment timeseries and drought sensitivity*** across trees and species in the Huai Kha Khaeng ForestGEO plot. a) Median of annualised growth increments across species (colours) and the whole plot (black) from 2009 - 2022 from late wet/early dry season dendroband measurements. 2010 and 2015 were moderate and very strong ENSO events respectively. b) Distribution of drought sensitivity of tree growth across all individuals in the two drought years. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/11](display/Fig2.png)

@@

We identified drought years of interest in the dendroband timeseries - 2010 and 2015 - based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses. 
Within the dendroband timeseries, two ENSO events had occurred; 2010 was a moderate ENSO event while 2015 was a very strong ENSO event [@REF]. 
<!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected. -->
We also calculate Standardised Precipitation-Evapotranspiration Index (SPEI) at the monthly scale using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
Defining drought as periods with two or more consecutive months with SPEI < -1 also resulted in identification of 2010 and 2015 as drought events. 
These years also corresponded with expert knowledge of major droughts that affected the region as well as meteorological records (see next paragraph; Fig. `r droughts`) and plot-level growth declines in the timeseries (Fig. `r drought_responses`). 
<!-- We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.  -->
<!-- Moreover, broad definitions based on climate variables (often not measured at the location), may not capture ecological realities within the plot.  -->
We note that although tree growth was unusally low in 2020, this year did not emerge as a drought year either in the examination of drought indicators or from expert knowledge.

@@

The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r droughts`).
In 2010, drought characteristics of lower precipitation, higher number of dry days, higher maximum temperature, higher VPD than the long-term mean and SPEI values below -1 occurred between February and April, in the dry season. 
On the other hand, in 2015, similar deviations occurred in May and June, in the wet season. 
In 2015, SPEI in May was -2, an anomalously low value that was not observed in any month in the entire dendroband census time period, suggesting a severe drought.

@@

The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig 2a). 
Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual, na.rm=T),2)` cm. 
Median annual increment differed by species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` cm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` cm) among the `r nrow(tree.time.sp)` species analysed.

@@

Trees on average had negative drought sensitivities for both droughts with considerable variation around it (Fig 2b). 
Median drought sensitivity across all trees in 2010 was `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)` while in 2015 it was `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`. 
In 2010, *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* and *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/2]`* had the lowest and highest median sensitivity of `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` and `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/2]` respectively. 
In 2015, *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* and *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/2]`* had the lowest and highest median sensitivity of `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` and `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/2]` respectively.

![***Figure 3: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010 and 2015. a) Relationship of the mean species sensitivities from model fits of an intercept-only model, with mean species deciduousness values. Line shown for significant correlation. b) Predicted sensitivity values across the observed range of deciduousness and topographic wetness index from a model with TWI, deciduosness and their interaction across all trees modelled for each drought year separately. For comments: https://github.com/forestgeo/growth-precip-thailand/issues/12](display/Fig3.png)

@@

Deciduousness was associated with lower negative sensitivities in 2015, but not in 2010 (Fig 3a). 
Species median sensitivities in 2010 and 2015 were not correlated with each other (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S5) but on average had lower mean and median sensitivities in 2015 than in 2010 (Fig S5), in alignment with median sensitivities across all trees. 
In 2015, predicted sensitivities from intercept-only models of sensitivity with a species random effect on the intercept showed a significant negative association with deciduousness value (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`). 
On the other hand, in 2010, predicted sensitivities had no significant association with deciduousness (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`). 
Moreover, models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`) but not in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).

@@

The effect of wetness on sensitivity was different in 2010 and 2015, with 2015 effects interacting with deciduousness. 
Individual tree responses to drought were not correlated with each other (Fig S6). 
Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`). 
The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
Combined, predicted sensitivities of deciduous trees was more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig 3b).

The opposite effects of TWI on sensitivity in the wet and dry season drought, and the influence of deciduousness was also confirmed by models that controlled for crown exposure and size as well as species variation in intercept and slopes (Fig 4a). 
These models showed small negative sensitivities in wetter sites in the dry season drought (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but strong positive effect of wetness in the wet season drought (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`). 
The positive effect of TWI in 2015 was, however, mainly due to the response of evergreen species; deciduousness had a significant negative correlation with the slopes of the TWI effect in 2015 (Fig 4b). 
Predictions across the range of observed variables also showed that deciduous species generally had larger negative sensitivities in 2015, which did not show increase with wetness (Fig 4c). 
On the other hand, neither the mean sensitivities across species nor their response to TWI was associated with deciduousness in 2010.

![***Figure 4: Drivers of variation in sensitivity among individuals***. All panels represent model results from models (one for each year) predicting sensitivity with microenvironment conditions with species random effect on all slopes. a) Coefficient plots showing median effects and 90% CI for wetness, exposure and DBH of tree. b) wetness effect across all species and its relationship with deciduousness value. Line represents a significant correlation. c) Predicted relationship of drought sensitivity with topographic wetness index across all species derived from model predictions. Black line represents overall relationship; coloured lines for each species reflects deciduousness values. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/13](display/Fig4.png)

@@

Crown exposure was associated with stronger negative sensitivities in both droughts.
In the same models with species random effects on intercept and slope, the simplex parameter for the ordered factor crown exposure in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 the negative effect was not as strong (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`). 
Predicted effects of crown exposure on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all crown exposure categories had negative sensitivities in 2015 (Fig 5b).

Accounting for effects of exposure, remaining effects of size on sensitivity differed between the two droughts. 
DBH was a significant predictor of exposure category, with larger DBH associated with higher exposure categories in 2010 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`) and 2015 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) (Fig 5a). 
Controlling for this direct effect of exposure, the remaining effect of size *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig 4a, Fig 5a).

![***Figure 5: Effects of size, exposure and water availability*** from combined models. Hypothesised relationships between microenvironment variables and drought sensitivity with their modelled effects in the HKK ForestGEO plot in the droughts of a) 2010 and b) 2015. Solid lines represent relationships where 90% credible intervals do not overlap 0, blue and red lines represent positive and negative effects respectively and line thickness is scaled to the effect size. c) Modelled sensitivity across crown exposure classes in 2010 and 2015. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/14](display/Fig5.png)

@@

<!---CONCLUSIONS---->
We show strong evidence that tropical tree drought sensitivity can be jointly and interactively shaped by drought characteristics, species adaptations and local environmental conditions (Table 1).
Analysing two droughts in a 14-year dendrometer band timeseries, we found that tropical tree species and individuals grew less than average in drought years, reflecting in lower plot-level growth (Fig 2). 
Comparing across droughts, we found that despite the *shorter duration of drought conditions*<!--careful here - didn't this continue into 2016?--> and greater absolute extremes of temperature, vapor pressure deficit, and lack of precipitation (Fig. 2a), the wet-season drought elicited stronger responses across trees, especially deciduous species that are expected to have stronger seasonality in growth phenology (Fig 3). 
Interestingly, in the wet-season drought, we also show that evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites (Fig 3b, Fig 4b). 
Further, besides consistent negative effects of crown exposure, we found that the drivers of variation among species and individuals in drought sensitivity were dissimilar between the drought events (Fig 4a). 
Water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts - larger trees (with likely deeper roots) in the dry-season drought and trees in wetter sites in the wet-season drought suffered smaller declines in growth (Fig 5). 
Taken together, our analysis of tree growth in two different drought years, and spanning meaningful variation, shows that species and individuals can respond differently to different drought events. 
Despite stand-level growth declines in drought years, we show that these broad patterns belie enormous heterogeneity in individual and species responses to drought in tropical forests with possible consequences for the maintenance of species diveristy and ecosystem stability.

>>>>

## Figures
fig_droughts <- 1
fig_drought_responses <- 2
fig_species <- 3
fig_microenvs <- 4
fig_dag <- 5

@@

![***Figure `r fig_droughts`. Climatic characteristics of drought years and average years*** in the Huai Kha Khaeng ForestGEO plot. For four daily climate variables dry days, precipitation, maximum temperature and VPD, panel a) represents rolling means for 30 days around the focal day for that year. Grey line represents the mean of this rolling mean from 2008 to 2019 and ribbons represent standard deviation around this mean. Panel b) represents anomalies of this rolling mean from the values across the timeseries. Values, therefore represents the number of standard deviations from the mean on a given day. c) shows Standardised Precipitation Evapotranspiration Index (SPEI) values for each month in 2010 and 2015. Across all panels, light blue background represents the average wet season, from May to October. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1.png)

@@

![***Figure `r fig_drought_responses`: Growth increment timeseries and drought sensitivity*** across trees and species in the Huai Kha Khaeng ForestGEO plot. a) Median of annualised growth increments across species (colours) and the whole plot (black) from 2009 - 2022 from late wet/early dry season dendroband measurements. 2010 and 2015 were moderate and very strong ENSO events respectively. b) Distribution of drought sensitivity of tree growth across all individuals in the two drought years. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/11](display/Fig2.png)

@@

We identified drought years of interest in the dendroband timeseries - 2010 and 2015 - based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses. 
Within the dendroband timeseries, two ENSO events had occurred; 2010 was a moderate ENSO event while 2015 was a very strong ENSO event [@REF]. 
<!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected. -->
We also calculate Standardised Precipitation-Evapotranspiration Index (SPEI) at the monthly scale using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
Defining drought as periods with two or more consecutive months with SPEI < -1 also resulted in identification of 2010 and 2015 as drought events. 
These years also corresponded with expert knowledge of major droughts that affected the region as well as meteorological records (see next paragraph; Fig. `r fig_droughts`) and plot-level growth declines in the timeseries (Fig. `r fig_drought_responses`). 
<!-- We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.  -->
<!-- Moreover, broad definitions based on climate variables (often not measured at the location), may not capture ecological realities within the plot.  -->
We note that although tree growth was unusally low in 2020, this year did not emerge as a drought year either in the examination of drought indicators or from expert knowledge.

@@

The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r fig_droughts`).
In 2010, drought characteristics of lower precipitation, higher number of dry days, higher maximum temperature, higher VPD than the long-term mean and SPEI values below -1 occurred between February and April, in the dry season. 
On the other hand, in 2015, similar deviations occurred in May and June, in the wet season. 
In 2015, SPEI in May was -2, an anomalously low value that was not observed in any month in the entire dendroband census time period, suggesting a severe drought.

@@

The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a). 
Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual, na.rm=T),2)` cm. 
Median annual increment differed by species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` cm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` cm) among the `r nrow(tree.time.sp)` species analysed.

@@

Trees on average had negative drought sensitivities for both droughts with considerable variation around it (Fig `r fig_drought_responses`b). 
Median drought sensitivity across all trees in 2010 was `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)` while in 2015 it was `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`. 
In 2010, *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* and *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/2]`* had the lowest and highest median sensitivity of `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` and `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/2]` respectively. 
In 2015, *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* and *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/2]`* had the lowest and highest median sensitivity of `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` and `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/2]` respectively.

![***Figure `r fig_species`: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010 and 2015. a) Relationship of the mean species sensitivities from model fits of an intercept-only model, with mean species deciduousness values. Line shown for significant correlation. b) Predicted sensitivity values across the observed range of deciduousness and topographic wetness index from a model with TWI, deciduosness and their interaction across all trees modelled for each drought year separately. For comments: https://github.com/forestgeo/growth-precip-thailand/issues/12](display/Fig3.png)

@@

Deciduousness was associated with lower negative sensitivities in 2015, but not in 2010 (Fig `r fig_species`a). 
Species median sensitivities in 2010 and 2015 were not correlated with each other (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S5) but on average had lower mean and median sensitivities in 2015 than in 2010 (Fig S5), in alignment with median sensitivities across all trees. 
In 2015, predicted sensitivities from intercept-only models of sensitivity with a species random effect on the intercept showed a significant negative association with deciduousness value (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`). 
On the other hand, in 2010, predicted sensitivities had no significant association with deciduousness (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`). 
Moreover, models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`) but not in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).

@@

The effect of wetness on sensitivity was different in 2010 and 2015, with 2015 effects interacting with deciduousness. 
Individual tree responses to drought were not correlated with each other (Fig S6). 
Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`). 
The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
Combined, predicted sensitivities of deciduous trees was more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig `r fig_species`b).

The opposite effects of TWI on sensitivity in the wet and dry season drought, and the influence of deciduousness was also confirmed by models that controlled for crown exposure and size as well as species variation in intercept and slopes (Fig `r fig_microenvs`a). 
These models showed small negative sensitivities in wetter sites in the dry season drought (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but strong positive effect of wetness in the wet season drought (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`). 
The positive effect of TWI in 2015 was, however, mainly due to the response of evergreen species; deciduousness had a significant negative correlation with the slopes of the TWI effect in 2015 (Fig `r fig_microenvs`b). 
Predictions across the range of observed variables also showed that deciduous species generally had larger negative sensitivities in 2015, which did not show increase with wetness (Fig `r fig_microenvs`c). 
On the other hand, neither the mean sensitivities across species nor their response to TWI was associated with deciduousness in 2010.

![***Figure `r fig_microenvs`: Drivers of variation in sensitivity among individuals***. All panels represent model results from models (one for each year) predicting sensitivity with microenvironment conditions with species random effect on all slopes. a) Coefficient plots showing median effects and 90% CI for wetness, exposure and DBH of tree. b) wetness effect across all species and its relationship with deciduousness value. Line represents a significant correlation. c) Predicted relationship of drought sensitivity with topographic wetness index across all species derived from model predictions. Black line represents overall relationship; coloured lines for each species reflects deciduousness values. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/13](display/Fig4.png)

@@

Crown exposure was associated with stronger negative sensitivities in both droughts.
In the same models with species random effects on intercept and slope, the simplex parameter for the ordered factor crown exposure in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 the negative effect was not as strong (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`). 
Predicted effects of crown exposure on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all crown exposure categories had negative sensitivities in 2015 (Fig `r fig_dag`b).

Accounting for effects of exposure, remaining effects of size on sensitivity differed between the two droughts. 
DBH was a significant predictor of exposure category, with larger DBH associated with higher exposure categories in 2010 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`) and 2015 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) (Fig 5a). 
Controlling for this direct effect of exposure, the remaining effect of size *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig `r fig_microenvs`a, Fig `r fig_dag`a).

![***Figure `r fig_dag`: Effects of size, exposure and water availability*** from combined models. Hypothesised relationships between microenvironment variables and drought sensitivity with their modelled effects in the HKK ForestGEO plot in the droughts of a) 2010 and b) 2015. Solid lines represent relationships where 90% credible intervals do not overlap 0, blue and red lines represent positive and negative effects respectively and line thickness is scaled to the effect size. c) Modelled sensitivity across crown exposure classes in 2010 and 2015. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/14](display/Fig5.png)

@@

<!---CONCLUSIONS---->
We show strong evidence that tropical tree drought sensitivity can be jointly and interactively shaped by drought characteristics, species adaptations and local environmental conditions (Table 1).
Analysing two droughts in a 14-year dendrometer band timeseries, we found that tropical tree species and individuals grew less than average in drought years, reflecting in lower plot-level growth (Fig `r fig_drought_responses`). 
Comparing across droughts, we found that despite the *shorter duration of drought conditions*<!--careful here - didn't this continue into 2016?--> and greater absolute extremes of temperature, vapor pressure deficit, and lack of precipitation (Fig. 2a), the wet-season drought elicited stronger responses across trees, especially deciduous species that are expected to have stronger seasonality in growth phenology (Fig `r fig_species`). 
Interestingly, in the wet-season drought, we also show that evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites (Fig `r fig_species`b, Fig `r fig_microenvs`b). 
Further, besides consistent negative effects of crown exposure, we found that the drivers of variation among species and individuals in drought sensitivity were dissimilar between the drought events (Fig `r fig_microenvs`a). 
Water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts - larger trees (with likely deeper roots) in the dry-season drought and trees in wetter sites in the wet-season drought suffered smaller declines in growth (Fig `r fig_dag`). 
Taken together, our analysis of tree growth in two different drought years, and spanning meaningful variation, shows that species and individuals can respond differently to different drought events. 
Despite stand-level growth declines in drought years, we show that these broad patterns belie enormous heterogeneity in individual and species responses to drought in tropical forests with possible consequences for the maintenance of species diveristy and ecosystem stability.
